True versus apparent shapes of bow shocks

### 2018-02-26 papers/Papers1234/app-rcurv-empirical.tex
Add material to appendix E on the circle fits to data

Consider a set of \(N\) points on the plane of the sky, with Cartesian
coordinates \(\bm{r}_k = (x_k, y_k)\) for \(k = 1 \dots N\).  We wish
to estimate the radius of curvature of the smooth curve that the set
of points is presumed to be sampled from.  To do this, we fit a circle
to the points as follows.  The circle is defined by its center,
\(\bm{r}_{\C} = (x_{\C}, y_{\C})\), and radius, \(R_{\C}\).  For a
given circle, we define a mean radius of the set of points from the
circle center:
\begin{equation}
  \label{eq:rcurv-mean-radius}
  \bar{R}(x_{\C}, y_{\C}) = \frac{1}{N}  \sum_{k = 1}^{N} \Abs{\bm{r}_k - \bm{r}_{\C}}  \ .
\end{equation}
We then optimize\footnote{%
  For example, using a Levenberg-Marquardt algorithm, such as
  \texttt{astropy.modeling.fitting.LevMarLSQFitter} from the AstroPy
  library \citep{Astropy-Collaboration:2013a}.} %
to find best-fit values \((x_{\C}^*, y_{\C}^*)\), which minimize the
objective function
\begin{equation}
  \label{eq:rcurv-objective-function}
  f(x_{\C}, y_{\C}) = \sum_{k = 1}^{N} \left(
    \abs{\bm{r}_k - \bm{r}_{\C}}  - \bar{R} (x_{\C}, y_{\C}) \right)^2 \ .
\end{equation}
The best-fit radius of curvature is then given by
\(R_{\C}^* = \bar{R}(x_{\C}^*, y_{\C}^*)\).

If we also know the position, \(\bm{r}_0 = (x_0, y_0)\), of the bow's
central source, then we can find the unit vector in the direction of
the bow's projected axis as
\begin{equation}
  \label{eq:rcurv-axis-vector}
  \uvec{\xi} = \frac {\bm{r}_0 - \bm{r}_{\C}^*} { \abs{\bm{r}_0 - \bm{r}_{\C}^*}} \ , 
\end{equation}
and the apex distance from the source as\footnote{%
  This is only valid if the resultant \(R_0 < R_{\C}^*\), otherwise
  the opposite sign of \(\uvec{\xi}\) must be taken.}
\begin{equation}
  \label{eq:rcurv-apex-distance}
  R_0 = \Abs{\bm{r}_{\C}^* + R_{\C}^* \uvec{\xi} - \bm{r}_0} \ .
\end{equation}
A refinement of the method is then to repeat the circle fit after
restricting the set of points to those lying within a certain angle
\(\Delta\theta\) of the bow axis by requiring
\begin{equation}
  \label{eq:rcurv-delta-theta}
  \uvec{\xi} \cdot 
  {\left( \bm{r}_k - \bm{r}_{\C}^* \right) }
  \ge   {\Abs{\bm{r}_k - \bm{r}_{\C}^*}}
  \cos \Delta\theta \ ,
\end{equation}
where we find that best results are obtained with \(\Delta\theta \le \ang{30}\).

If quantitative estimates exist for the uncertainties, \(\epsilon_k\), in the
measurements of \(\bm{r}_k\), then it is appropriate to incorporate
weights of \(\epsilon_k^{-2}\) in the objective function.  In cases where the
bow shape is traced by eye, based on real or synthetic observations, a
more practical approach is to maintain uniform weighting but to place
a greater density of points \(\bm{r}_k\) in regions where the shape is
well-determined and to place them more sparsely in regions where the
shape is less certain.

Note that this method is not necessary if the orientation of the bow
axis is known a~priori, in which case the Taylor series method
described in \S~\ref{sec:plan-alat-bow} is more efficient and
accurate.

>>>>

Consider a set of \(N\) points on the plane of the sky, with Cartesian
coordinates \(\bm{r}_k = (x_k, y_k)\) for \(k = 1 \dots N\).  We wish
to estimate the radius of curvature of the smooth curve that the set
of points is presumed to be sampled from.  To do this, we fit a circle
to the points as follows.  The circle is defined by its center,
\(\bm{r}_{\C} = (x_{\C}, y_{\C})\), and radius, \(R_{\C}\).  For a
given circle, we define a mean radius of the set of points from the
circle center:
\begin{equation}
  \label{eq:rcurv-mean-radius}
  \bar{R}_\C(x_{\C}, y_{\C}) = \frac{1}{N}  \sum_{k = 1}^{N} \Abs{\bm{r}_k - \bm{r}_{\C}}  \ .
\end{equation}
We then optimize
% \footnote{%
%   For example, using a Levenberg-Marquardt algorithm, such as
%   \texttt{astropy.modeling.fitting.LevMarLSQFitter} from the AstroPy
%   library \citep{Astropy-Collaboration:2013a}.} %
to find best-fit values \((x_{\C}^*, y_{\C}^*)\), which minimize the
objective function
\begin{equation}
  \label{eq:rcurv-objective-function}
  f(x_{\C}, y_{\C}) = \sum_{k = 1}^{N} \left(
    \abs{\bm{r}_k - \bm{r}_{\C}}  - \bar{R}_\C (x_{\C}, y_{\C}) \right)^2 \ .
\end{equation}
The best-fit radius of curvature is then given by
\(R_{\C}^* = \bar{R}_\C(x_{\C}^*, y_{\C}^*)\).

If we also know the position, \(\bm{r}_0 = (x_0, y_0)\), of the bow's
central source, then we can find the unit vector in the direction of
the bow's projected axis as
\begin{equation}
  \label{eq:rcurv-axis-vector}
  \uvec{\xi} = \frac {\bm{r}_0 - \bm{r}_{\C}^*} { \abs{\bm{r}_0 - \bm{r}_{\C}^*}} \ , 
\end{equation}
and the apex distance from the source as\footnote{%
  This is only valid if the resultant \(R_0 < R_{\C}^*\), otherwise
  the opposite sign of \(\uvec{\xi}\) must be taken.}
\begin{equation}
  \label{eq:rcurv-apex-distance}
  R_0 = \Abs{\bm{r}_{\C}^* + R_{\C}^* \uvec{\xi} - \bm{r}_0} \ .
\end{equation}
A refinement of the method is then to iteratively repeat the circle
fit after restricting the set of points to those lying within a
certain angle \(\Delta\theta\) of the bow axis, where we find that best results
are obtained with \(\Delta\theta \approx \ang{60}\) to \ang{75}.  That is, 
\begin{equation}
  \label{eq:rcurv-delta-theta}
  \Abs{\theta_k} < \Delta\theta \ , 
\end{equation}
where the angle \(\theta_k\) of each point from the axis, measured at the
source position \(r_0\), can be calculated\footnote{%
  Note that the cross product (\(\times\)) of the 2d vectors is assumed to
  yield the (signed scalar) \(z\) component, also known as the
  ``perpendicular dot product''.} %
as
\begin{equation}
  \label{eq:rcurv-theta-k}
  \theta_k = \arctan \left[  
    \frac{\left( \bm{r}_k - \bm{r}_0 \right) \times \uvec{\xi}}
    {\left( \bm{r}_k - \bm{r}_0 \right) \cdot \uvec{\xi}}
  \right] \ .
\end{equation}
Two or three iterations are sufficient for convergence in most cases,
although in some cases it is possible that the process will converge
to a stable flip-flop oscillation between two different solutions.
This is due to the dependence of the \(\theta_k\), via \(\uvec{\xi}\), on
the \(\bm{r}_{\C}^*\) of the previous iteration, which can lead to
points entering and leaving the fitted set.  We have not found this to
be a serious problem in practice, since the two solutions tend to be
very close to one another.  It could be mitigated by averaging
\(\bm{r}_{\C}^*\) over two previous iterations.  The alternative of
measuring the angle with respect to the center of curvature,
\(\bm{r}_{\C}^*\), instead of the source, \(\bm{r}_0\), is found to be
much less stable.

If quantitative estimates exist for the uncertainties, \(\epsilon_k\), in the
measurements of \(\bm{r}_k\), then it is appropriate to incorporate
weights of \(\epsilon_k^{-2}\) in the objective function.  However, in cases
where the bow shape is traced by eye, based on real or synthetic
observations, a more practical approach is to maintain uniform
weighting but to place a greater density of points \(\bm{r}_k\) in
regions where the shape is well-determined and to place them more
sparsely in regions where the shape is less certain.

Since there is no guarantee of symmetry about the axis \(\uvec{\xi}\),
the perpendicular radius will in general be different in the two wings
of the bow, with values \(R_{90}^+\) and \(R_{90}^-\).  These can be
estimated by defining
\begin{equation}
  \label{eq:rcurv-R-k}
  R_k = \Abs{\bm{r}_k - \bm{r}_0}
\end{equation}
and linearly interpolating between the points \((\theta_k, R_k)\) at
\(\theta = \pm \ang{90}\).

Our Python language implementation of this algorithm is freely
available at \url{https://github.com/div-B-equals-0/circle-fit}.  An
example application to real data is given in \S~\ref{sec:obs} and
Figure~\ref{fig:000-400-fit}.  Note that this method is not necessary
if the orientation of the bow axis is known a~priori, in which case
the Taylor series method described in \S~\ref{sec:plan-alat-bow} is
more efficient and accurate.
